---
title: "Cohort construction with provenance"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cohort construction with provenance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`cohort` provides the `CohortPipeline` R6 class for building analytic
cohorts with full provenance. Cohort construction is kept strictly upstream
of analysis: the pipeline produces analytic data tables, then hands them
off — typically into a `plnr::Plan`.

## A worked example

We start with a small simulated dataset.

```{r}
library(cohort)
library(data.table)

set.seed(1)
d <- data.table(
  id  = 1:200,
  age = sample(c(NA, 16:80), 200, replace = TRUE),
  sex = sample(c("F", "M", NA), 200, replace = TRUE,
               prob = c(0.48, 0.48, 0.04)),
  grp = sample(c("a", "b"), 200, replace = TRUE)
)
```

### Step 1: install a base table

```{r}
cp <- CohortPipeline$new()
cp$load(d)
```

`CohortPipeline$new()` makes a defensive copy of `d` once. The user's
data table is never mutated, no matter what subsequent operations you
run on the pipeline.

### Step 2: apply root-level exclusions

Every exclusion takes a human-readable reason and an R expression as a
string. The string is parsed and evaluated against the rows currently
included on the branch.

```{r}
cp$exclude_and_track("root", "Missing sex", "is.na(sex)")
cp$exclude_and_track("root", "Missing age", "is.na(age)")
cp$exclude_and_track("root", "Under 18",    "age < 18")
```

`NA` predicate results are treated as `FALSE` (rows are kept) so you
can write predicates without defensive `!is.na(...)` clauses for every
column.

### Step 3: branch into sub-cohorts

```{r}
cp$new_cohort("adults_female", from = "root")
cp$exclude_and_track("adults_female", "Not female", "sex != 'F'")

cp$new_cohort("adults_grp_a", from = "root")
cp$exclude_and_track("adults_grp_a", "Group not a", "grp != 'a'")
```

A branch starts identical to its parent at the moment of branching.
Subsequent exclusions on the parent never propagate to a child; sibling
branches evolve independently.

### Step 4: derive cached artifacts

`set_artifact()` is for reusable per-cohort objects (analytic-ready data
tables, summary statistics, baseline tables). Each callback receives a
copy of the included rows and the named list of artifacts already
attached to the cohort.

```{r}
cp$set_artifact("dt_for_analysis",
  from = "adults_female",
  fn = function(dt, sib) {
    dt[, age_group := cut(age,
      breaks = c(18, 30, 50, Inf),
      right = FALSE,
      labels = c("18-29", "30-49", "50+"))]
    dt
  }
)
```

Each `set_artifact()` callback receives a *fresh* copy of the cohort's
included rows — it does not see modifications made by previous
artifacts. To chain derivations, read the previous artifact off the
`sib` argument:

```{r}
cp$set_artifact("baseline_table",
  from = "adults_female",
  fn = function(dt, sib) {
    sib$dt_for_analysis[, .(.N, mean_age = mean(age)),
      by = age_group][order(age_group)]
  }
)
```

### Step 5: inspect the cohort tree

```{r}
print(cp)
cp$list_cohorts()
cp$consort()
```

`$consort()` returns a long-form table — one row per exclusion step,
across every branch. Each branch contributes only its own steps; steps
inherited from a parent at branch time are reported under the parent.

### Step 6: hand off to plnr

The bulk-ingest helper `add_data_from_cohort()` registers a cohort's
artifacts on a `plnr::Plan` as named data entries. Each entry is named
`paste0(prefix, cohort, "__", artifact)`.

```{r, eval = requireNamespace("plnr", quietly = TRUE)}
plan <- plnr::Plan$new()
add_data_from_cohort(plan, cp, cohort = "adults_female")
plan$get_data()
```

You can now `add_argset()` and `add_analysis()` on the plan as usual.

## Schemas

Use schemas to declare a column-type contract on a branch and verify it
before downstream code consumes the data.

```{r}
cp$declare_schema("adults_female", schema = list(
  age = list(type = "numeric",   na = FALSE),
  sex = list(type = "character", na = FALSE)
))
cp$validate()
```

If a column is missing, has the wrong type, or carries unexpected `NA`s,
`$validate()` throws a single error listing every problem. Pass
`auto_validate = TRUE` to `CohortPipeline$new()` to fail at the failure
site (every `$new_cohort()` and `$set_artifact()` call validates after).

## CONSORT diagrams

`$draw_consort_panels()` renders one or more CONSORT panels using the
`consort` package. Each panel walks a sequence of cohort names; their
exclusion steps become bullet blocks on the spine.

```r
cp$draw_consort_panels(
  panels = list(
    "Primary analytic chain" = c(
      "All participants" = "root",
      "Adult females"    = "adults_female"
    )
  ),
  file = "Figure_1_CONSORT.pdf"
)
```

## Mutation contract (summary)

| Method                                | Returns                      | Mutation safe?       |
|---------------------------------------|------------------------------|----------------------|
| `$get_included(c)` / `copy = TRUE`    | Independent copy             | Yes                  |
| `$get_included(c, copy = FALSE)`      | Row-subset of base           | **No** — read-only   |
| Callback in `$set_artifact`           | Independent copy of subset   | Yes                  |
| `$get_everyone(c)`                    | Independent copy + status    | Yes                  |

Operations that go through the public API never modify the user's input
data table or another branch.

## Performance notes

- Branching is O(n) in the number of rows of the base table. The data
  values themselves are stored once and shared across the tree.
- `$exclude_and_track()` evaluates the predicate against the included
  subset only; predicates may safely assume earlier exclusions have
  already removed invalid rows.
- The exclusion log is accumulated as a list and materialized as a
  `data.table` only on read (`$consort()`, `$list_cohorts()`), avoiding
  quadratic `rbind` growth.
