---
title: "Cohort construction with provenance"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cohort construction with provenance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`cohort` provides the `CohortPipeline` R6 class for building analytic
cohorts with full provenance. Cohort construction is kept strictly upstream
of analysis: the pipeline produces analytic data tables, then hands them
off to whatever consumes them downstream.

## A worked example

We start with a small simulated dataset.

```{r}
library(cohort)
library(data.table)

set.seed(1)
d <- data.table(
  id  = 1:200,
  age = sample(c(NA, 16:80), 200, replace = TRUE),
  sex = sample(c("F", "M", NA), 200, replace = TRUE,
               prob = c(0.48, 0.48, 0.04)),
  grp = sample(c("a", "b"), 200, replace = TRUE)
)
```

### Step 1: install a base table

```{r}
cp <- CohortPipeline$new()
cp$load(d)
```

`CohortPipeline$new()` makes a defensive copy of `d` once. The user's
data table is never mutated, no matter what subsequent operations you
run on the pipeline.

### Step 2: apply root-level exclusions

Every exclusion takes a human-readable reason and an R expression as a
string. The string is parsed and evaluated against the rows currently
included on the branch.

```{r}
cp$exclude_and_track("root", "Missing sex", "is.na(sex)")
cp$exclude_and_track("root", "Missing age", "is.na(age)")
cp$exclude_and_track("root", "Under 18",    "age < 18")
```

`NA` predicate results are treated as `FALSE` (rows are kept) so you
can write predicates without defensive `!is.na(...)` clauses for every
column.

### Step 3: branch into sub-cohorts

```{r}
cp$new_cohort("adults_female", from = "root")
cp$exclude_and_track("adults_female", "Not female", "sex != 'F'")

cp$new_cohort("adults_grp_a", from = "root")
cp$exclude_and_track("adults_grp_a", "Group not a", "grp != 'a'")
```

A branch starts identical to its parent at the moment of branching.
Sibling branches evolve independently.

### The freeze rule

A cohort becomes **frozen** the first time either:

1. another cohort branches from it, or
2. an artifact is set on it.

After freezing, `$exclude_and_track()` on that cohort errors. The rule
guarantees that a cohort's name maps to one definition forever: once
children depend on it, its exclusion list is fixed. The practical
workflow is "apply all exclusions on a cohort, then branch from it or
attach artifacts." Multi-way forks are unaffected — you can branch a
frozen cohort as many times as you like.

### Step 4: derive cached artifacts

`set_artifact()` is for reusable per-cohort objects (analytic-ready data
tables, summary statistics, baseline tables). Each callback receives a
copy of the included rows and the named list of artifacts already
attached to the cohort.

```{r}
cp$set_artifact("dt_for_analysis",
  from = "adults_female",
  fn = function(dt, sib) {
    dt[, age_group := cut(age,
      breaks = c(18, 30, 50, Inf),
      right = FALSE,
      labels = c("18-29", "30-49", "50+"))]
    dt
  }
)
```

Each `set_artifact()` callback receives a *fresh* copy of the cohort's
included rows — it does not see modifications made by previous
artifacts. To chain derivations, read the previous artifact off the
`sib` argument:

```{r}
cp$set_artifact("baseline_table",
  from = "adults_female",
  fn = function(dt, sib) {
    sib$dt_for_analysis[, .(.N, mean_age = mean(age)),
      by = age_group][order(age_group)]
  }
)
```

### Step 5: inspect the cohort tree

```{r}
print(cp)
cp$list_cohorts()
cp$consort()
```

`$consort()` returns a long-form table — one row per exclusion step,
across every branch. Each branch contributes only its own steps; steps
inherited from a parent at branch time are reported under the parent.

### Step 6: hand off downstream

Cached artifacts are plain R objects — pull them out with
`$get_artifact()` and pass them to whatever consumes the analytic data.

```{r}
analytic_dt <- cp$get_artifact("adults_female", "dt_for_analysis")
baseline    <- cp$get_artifact("adults_female", "baseline_table")

head(analytic_dt)
baseline
```

If you use an analysis-orchestration package, register each artifact as
a named data entry — typically one short loop over `cp$list_artifacts()`.

## Schemas

Use schemas to declare a column-type contract on a branch and verify it
before downstream code consumes the data.

```{r}
cp$declare_schema("adults_female", schema = list(
  age = list(type = "numeric",   na = FALSE),
  sex = list(type = "character", na = FALSE)
))
cp$validate()
```

If a column is missing, has the wrong type, or carries unexpected `NA`s,
`$validate()` throws a single error listing every problem. Pass
`auto_validate = TRUE` to `CohortPipeline$new()` to fail at the failure
site (every `$new_cohort()` and `$set_artifact()` call validates after).

## CONSORT diagrams

The simplest way to draw CONSORT diagrams is `$plot()`. With no
arguments it renders one panel per cohort, walking the root-to-cohort
path; cohort names become box labels.

```{r consort, fig.width = 10, fig.height = 6, fig.align = "center", out.width = "85%"}
cp$plot()
```

You can name specific cohorts:

```{r, fig.width = 7, fig.height = 6, fig.align = "center", out.width = "85%"}
cp$plot(c("adults_female", "adults_grp_a"))
```

Or save to disk:

```r
cp$plot(file = "Figure_1_CONSORT.pdf")
```

For full control over labels, panel titles, side branches and layout,
use `$draw_consort_panels()` (see `?CohortPipeline`).

## Incremental caching

For long-running scripts, supply a `cache_file` to `CohortPipeline$new()`
to enable incremental re-execution. On the first run, a snapshot of the
pipeline is written to disk; on subsequent runs, the snapshot is loaded
and each operation in the script is replayed against it. Operations
matching the cached log are no-ops (cursor advances, no work);
divergent operations truncate the log and recompute only the affected
cohort and its descendants.

```r
cp <- CohortPipeline$new(d, cache_file = "cache/cp.rds")
on.exit(cp$save(), add = TRUE)

cp$exclude_and_track("root", "Missing sex", "is.na(sex)")
cp$exclude_and_track("root", "Under 18",    "age < 18")
cp$new_cohort("adults_female", from = "root")
cp$exclude_and_track("adults_female", "Not female", "sex != 'F'")
```

Cache hits compare:

| Method                | Cache key                          |
|-----------------------|------------------------------------|
| `exclude_and_track`   | `(branch, reason, expr_str)`       |
| `new_cohort`          | `(name, from)`                     |
| `declare_schema`      | overwrites; not cached             |
| `set_artifact`        | `(name, from, body(fn), argset)`   |

For `set_artifact`, the 3-argument form `function(dt, sib, argset)` is
recommended when caching: the `argset` list carries the data
dependencies explicitly and participates in the cache key, so changing
e.g. `argset = list(washout = 84L)` to `washout = 90L` triggers a
recompute. The 2-argument form `function(dt, sib)` still works but
won't catch closure-captured changes.

The cache key uses `body(fn)` literally; if `fn` calls a helper that
you change, the cache cannot detect that. Either include a version tag
in `argset`, or call `cp$invalidate(cohort)` to drop the cohort and
force recompute.

## Mutation contract (summary)

| Method                       | Returns                    | Mutation safe? |
|------------------------------|----------------------------|----------------|
| `$get_included(c)`           | Independent copy           | Yes            |
| Callback in `$set_artifact`  | Independent copy of subset | Yes            |
| `$get_everyone(c)`           | Independent copy + status  | Yes            |

Operations that go through the public API never modify the user's input
data table or another branch.

## Performance notes

- Branching is O(n) in the number of rows of the base table. The data
  values themselves are stored once and shared across the tree.
- `$exclude_and_track()` evaluates the predicate against the included
  subset only; predicates may safely assume earlier exclusions have
  already removed invalid rows.
- The exclusion log is accumulated as a list and materialized as a
  `data.table` only on read (`$consort()`, `$list_cohorts()`), avoiding
  quadratic `rbind` growth.
